---
title: "<i>Etheostoma cragini</i> behavioral RNAseq pipeline"
author: "Coby McDonald"
date: "Last updated: `r Sys.Date()`"
output: 
  # html_document:
  #   df_print: paged
  #   toc: true
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE, eval=F, echo=F}
library(tidyverse)
library(here)
```

# Overview

- Project: Chris Kopack, 3rd PhD chapter
- Main questions: Does behavioral enrichment improve predator avoidance behavior in the Arkansas darter (<i>Etheostoma cragini</i>)? If so, what genes are involved in avoidance and learned avoidance?
- Approach: predator avoidance behavioral trial, followed by 3' RNA-seq of neural tissue

The main pipeline we'll use is:

1. QC: `FastQC`+ `MultiQC`
1. +/- read trimming: `trimgalore` and repeat QC
1. Alignment and quantification: `STAR`
1. Differential expression analysis: `DESeq2`
1. Functional enrichment testing: `emapper` and `topGO`

# Step 1: check read quality

__Programs:__ `FastQC` and `MultiQC`

__Explanation:__ First, we visually examine our sequencing results. `FastQC` generates standard QC metrics on a per file basis. `MultiQC` aggregates these results to make it easier to visually check hundreds of files at once.

__Code:__ [fastqc_multiqc_raw.sh](scripts/fastqc_multiqc_raw_slurm.sh)

__Results:__ See `MultiQC` report [here.](results_github/multiqc_report_raw.html)

# Step 2: trim low-quality reads, repeat QC

__Programs:__ `TrimGalore`, `FastQC`, `MultiQC`

__Explanation:__ For the purposes of counting applications (e.g. RNA-seq differential gene expression), read trimming is not required. This is because modern aligners (like `STAR`) perform local alignment and automatically soft-clip any non-matching sequences (e.g. adapter content).

Note: if you want to do transcriptome assembly, you *should* trim reads for both quality and adapter content.

__Code:__ [trimgalore.sh](scripts/trimgalore_slurm.sh)

__Results:__ See `MultiQC` report [here.](results_github/multiqc_report_trimmed.html)

Following trimming, read quality is marginally improved.

# Step 3: Read alignment and quantification

__Program:__ `STAR`

__Explanation:__ Provided you have a genome, `STAR` is by far the most commonly used RNA-seq aligner. It's super fast and has the added benefit of being able to perform quantification as well. `STAR` consists of two steps: first, generating genome indexes, then mapping reads to the indexed genome. There are a variety of parameters to fiddle with, but the few teleost RNA-seq papers I found tended to use default parameters.

__Code:__ [starquant_slurm_etheo.sh](scripts/starquant_slurm_etheo.sh)

__Results:__ Running `STAR` in quantification mode outputs a number of *ReadsPerGene.out.tab files. These output files have a row for each gene and four columns: gene ID, counts for unstranded RNA-seq, counts for 1st strand aligned to RNA, counts for 2nd strand aligned. I believe Chris did an unstranded library prep. Thus, we can extract the second column from each file and create a counts matrix.

This [counts matrix](results_github/ecrag_readcounts_raw.txt) can then be imported into `DESeq2.`

__Note: the % uniquely mapped reads for each sample is on the low side (~60-70%). Typically we shoot for 70-90%.__

To troubleshoot this, I subset 1000 unmapped sequences from one of the samples and blasted them against the non-redundant protein database. Sequences generally seem to align to: ribosomal RNA, mitochondrial RNA, and random genes in other fish species. This suggests to me that perhaps ribosomal depletion was not optimal during library prep, and that the <i>E. cragini</i> genome is incompletely annotated. Neither are huge issues.

We can confirm that this low-ish mapping rate isn't a big issue by re-running `STAR` with trimmed reads instead of raw reads. This increases our mapping percentages to be (mostly) ≥ 70%.


# Step 5: GO enrichment analyses

__Programs:__ `eggnog-mapper`, `topGO`

__Explanation:__ Although *E. cragini* has a published genome, it does not have gene ontology annotation. Thus, before we can perform GO enrichment, we must first retrieve all GO terms associated with our genes. We can do this with the functional annotation program `eggNOG-mapper`. `eggNOG-mapper` can take a fasta file of query sequences (either proteins or genes), run a Diamond blastp (or blastx) search, and carry out functional annotation for the queries with hits to eggNOG proteins.

I ran the emapper.py function against a taxonomy-restricted Diamond database, searching only Teleostei sequences. Restricting to a particular taxon makes the search much faster.

Following this, we can get the necessary GO ids to run `topGO`.

Basic steps:

1. Run emapper.py function (see [eggnog_annot.sh](scripts/eggnog_annot.sh))
1. Add GO IDs to gene IDs from *E. cragini* gtf file (see [add_annotations.R](scripts/add_annotations.R))
1. Run `topGO` (see [topGO_new.R](scripts/topGO_new.R))

